---
title: "Crash Course in R"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---


## Use a script file

Let's all make a new directory to work in today. 

Let's create a new script file in that directory.

Scripts make it easier to repeat your work. You can also add comments using the pound sign.

Short cut to execute commands and functions:

Mac:
[Command] + return

Windows
[Control] + return (RStudio)
[Control] + r (R gui)


### Setting your working directory

```{r, eval=FALSE}
setwd("[your dir name here]")
```


## R as a calculator

```{r}
5 + 3

8^3

6*3-1

# Order of operations follows PEMDAS
6*(3-1)
```

## Variables

We can save things within our session as **variables**

```{r}
pop_1 <- 1200
pop_2 <- 500

pop_total <- pop_1 + pop_2

pop_1 * 2

# Etc.
```

### Challange

I chagne pop_1

```{r}
pop_1 <- 2000
```

What is `pop_total` now?


## Loading Data

Best to use the full path to your data, but could also change into the directory you're data is in, then call it in there.


```{r}
fral_pres <- read.csv(file = "https://www.dropbox.com/s/x7s7fpu4bepj7xf/F_alnus_CompiledPres.csv?dl=1")
```


Let's have a look at these data

```{r}
head(fral_pres)

tail(fral_pres)

summary(fral_pres)

names(fral_pres)

str(fral_pres)
```


## Navigating data

What if we wanted to look at specific row/column entries?

```{r}
fral_pres[1, 1]
fral_pres[1, 2]
```

Let's get a specific row.

```{r}
fral_pres[3, ]
```

And a whole column

```{r, results='hide'}
fral_pres[ , 2 ]
```

Just part of the column

```{r}
fral_pres[1:10, 2 ]
```


Specific rows

```{r}
fral_pres[c(3, 5, 7), ]
```

How about rows that meet certain criteria?

```{r, results='hide'}
subset(fral_pres, FIELD_HERB == "Herbarium")
```



## Fixing or Cleaning Data

Let's say we realized that we had a mistake in our data. 
For example, one of `UNCERTAIN` values was recorded incorrectly. 
How can we change this?

```{r}
fral_pres_fixed <- fral_pres
fral_pres_fixed$UNCERTAIN[1]
fral_pres_fixed$UNCERTAIN[1] <- 20
fral_pres_fixed$UNCERTAIN[1]

```


## Subset the data

Let's get only a subset of these data, selecting from the `data.frame` by columns.

```{r}
fral_pres_subset <- fral_pres[c("SPEC", "LONG", "LAT")]
```

Next, let's rename our columns so they are in the format used in Wallace.

```{r}
names(fral_pres_subset) 
names(fral_pres_subset) <- c("name", "longitude", "latitude")
```




Let's now make a new file with the fixed data.

```{r}
write.csv(x = fral_pres_subset, file = "~/Dropbox/SCCS-Workshop/fral_pres.csv", row.names = FALSE)
```


## Simple calculations / built-in functions

Some statistics of note.

```{r}
mean(fral_pres$LAT)
max(fral_pres$LAT)
min(fral_pres$LAT)
median(fral_pres$LAT)
```


### Challenge

Use indexing and the functions we just learned to determin the mean, min, and max latitude and longitude of all of the Herbarium specimens vs. the Field specimens.


## Simple plots

```{r}
plot(x = fral_pres$LONG, y = fral_pres$LAT)
```


